\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-29 paper.qmd
Changed the abstract (worked with Thiyanga's comments also)

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional ($p\text{-}D$) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. To help evaluate the NLDR we have developed an algorithm to show the $2\text{-}D$ NLDR model in the $p\text{-}D$ space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which $2\text{-}D$ layout is the best representation of the $p\text{-}D$ distribution and see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data-space]
keywords-formatted: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data-space]

@@

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", position = c(0.92, 0.95))

@@

trimesh_two_curvy_removed1_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("b")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Best fit"
#| fig-width: 10
#| fig-height: 10

@@

tsne_two_curvy_clust <- tsne_two_curvy_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

## hexagon binning to have regular hexagons
hb_obj_tsne_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 10, 
  r2 = r2_tsne)

a1_2_tsne <- calc_bins_y(
  bin1 = 10, 
  r2 = r2_tsne
)$a1

@@

@@

nldr_two_curvy_original + tsne_two_curvy_clust +
  trimesh_two_curvy_removed1_with_data + trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_umap_model +
  two_curvy_proj_tsne_model +
  two_curvy_proj_umap_all_model +
  two_curvy_proj_tsne_all_model +
  plot_layout(ncol = 2)
```

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data ($p\text{-}D$) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. The NLDR methods and (hyper)parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the $p\text{-}D$ data, we have developed an algorithm to show the \gD{} NLDR model in the $p\text{-}D$ space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data vizualization, non-linear dimension reduction, tour]
keywords-formatted: [high-dimensional data vizualization, non-linear dimension reduction, tour]

@@

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a")

@@

trimesh_two_curvy_removed1_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  #xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("a2")

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Model in \\gD{}, on the layout and two views of the fit in projections from $7\\text{-}D$, for the two-curvy clusters data ($n =  2000$ and $p = 7$). The fitted model"
#| fig-width: 10
#| fig-height: 10

@@

tsne_two_curvy_clust_plt <- tsne_two_curvy_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("a", c(0.08, 0.93)) +
  theme(aspect.ratio = 1)

## hexagon binning to have regular hexagons
hb_obj_tsne_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 8, 
  r2 = r2_tsne)

a1_2_tsne <- calc_bins_y(
  bin1 = 8, 
  r2 = r2_tsne
)$a1

@@

benchmark_two_curvy4 <- 0.2

@@

nldr_two_curvy_original + tsne_two_curvy_clust_plt +
  trimesh_two_curvy_removed1_with_data + trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_umap_model +
  two_curvy_proj_tsne_model +
  two_curvy_proj_umap_all_model +
  two_curvy_proj_tsne_all_model +
  plot_layout(ncol = 2)
```

### 2024-09-29 paper.tex
Changed the abstract (worked with Thiyanga's comments also)

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension reduction, hexagon
binning, low-dimensional representation, tour, data
vizualization, model-in-the-data-space},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\bigskip
\bigskip
\begin{abstract}
Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional
(\(p\text{-}D\)) data using non-linear transformation. The methods and
parameter choices can create wildly different representations, making it
difficult to decide which is best, or whether any or all are accurate or
misleading. NLDR often exaggerates random patterns, sometimes due to the
samples observed. But NLDR views have an important role in data analysis
because, if done well, they provide a concise visual (and conceptual)
summary of \(p\text{-}D\) distributions. To help evaluate the NLDR we
have developed an algorithm to show the \(2\text{-}D\) NLDR model in the
\(p\text{-}D\) space, viewed with a tour. One can see if the model fits
everywhere or better in some subspaces, or completely mismatches the
data. It is used to evaluate which \(2\text{-}D\) layout is the best
representation of the \(p\text{-}D\) distribution and see how different
methods may have similar summaries or quirks.
\end{abstract}

\noindent%
{\it Keywords:} high-dimensional data, dimension reduction, hexagon
binning, low-dimensional representation, tour, data
vizualization, model-in-the-data-space
\vfill

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data vizualization, non-linear dimension
reduction, tour},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\bigskip
\bigskip
\begin{abstract}
Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional data
(\(p\text{-}D\)) by applying a non-linear transformation. NLDR often
exaggerates random patterns, sometimes due to the samples observed. But
NLDR views have an important role in data analysis because, if done
well, they provide a concise visual (and conceptual) summary of
\(p\text{-}D\) distributions. The NLDR methods and (hyper)parameter
choices can create wildly different representations, making it difficult
to decide which is best, or whether any or all are accurate or
misleading. To help assess the NLDR and decide on which, if any, is the
most reasonable representation of the structure(s) present in the
\(p\text{-}D\) data, we have developed an algorithm to show the \gD{}
NLDR model in the \(p\text{-}D\) space, viewed with a tour, a movie of
linear projections. From this, one can see if the model fits everywhere,
or better in some subspaces, or completely mismatches the data. Also, we
can see how different methods may have similar summaries or quirks.
\end{abstract}

\noindent%
{\it Keywords:} high-dimensional data vizualization, non-linear
dimension reduction, tour
\vfill

@@

\caption{Model in \gD{}, on the layout and two views of the fit in
projections from \(7\text{-}D\), for the two-curvy clusters data
(\(n =  2000\) and \(p = 7\)). The fitted model}
